Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-08 manu-draft.qmd
adding draft content to intro

-   Overview of metabolism - importance
-   Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
-   Applications to estuaries - challenges
-   Goals and objectives

@@

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`. Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.

>>>>

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere. As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011] . Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006] . Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$). As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, also referred to as net ecosystem production and net community production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.

Methods for estimating metabolic rates in aquatic environments are numerous . Kemp and Testa (2011) and Staehr et al. (2012) summarized the main approaches for quantifying ecosystem metabolism in aquatic systems (including incubations, the open-water method, ecosystem budgets, oxygen isotopes, and inert gases).

Complementing the incubation-based methods are open-water methods (Odum, 1956), which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (GPP, equals net primary production + autotrophic respiration), ecosystem respiration (ER, equals heterotrophic + autotrophic respiration), and NEP (equals GPP -- ER). A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries (Caffrey, 2004); a similar synthesis by Yvon-Durocher et al. (2012) was exclusively based on Caffrey (2004). As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Such a model should be informed by the available data at a given location and have the ability to incorporate prior information on key parameters if available. Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs. A potentially useful application developed for streams is the BAyesian Single-station Estimation method.

Application of open-water techniques for estimating metabolism is especially problematic based on the assumptions and poor error constraints of existing methods that rely on continuous measurements of dissolved oxygen. Fundamental to these challenges is the high temporal and spatial variability noted in global syntheses of primary production and respiration. Cloern et al. (2014) found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013).

The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production (P), ecosystem respiration (R), net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`. Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.
